---
title: "Workshop on Data Simulation & Power Analysis"
subtitle: ""
page-layout: full
editor_options: 
  chunk_output_type: console
---
```{r, include = FALSE}
install.packages('xfun')
library('xfun')
```

Welcome! This website will form the basis of the 2024 ManyBabies workshop on Data Simulation & Power Analysis. Before we start, I would like to emphasise that this workshop has grown out of discussion with lots of different people and is a true collaborative effort, which accords nicely with the general philosophy of ManyBabies projects. The approach to data simulation and power analysis explored here is closely associated with the data analysis team on the ManyBabies5 project (https://manybabies.org/MB5/). This series of meetings among researchers on the data analysis team was a fundamentally exploratory process being guided by a what-if mindset. 

For example, how do predictors with two or three levels impact the power to detect an effect on infant looking times? How does variability in individual labs come to affect generalisability and statistical power? What is the optimal balance between various pracical constraints (e.g., an upper bound on the number of stimulus items that infants can attend to) and statistical inference (e.g., how much of a decrease in power are we willing to accept based on the above constraints)? Can these results inform the experimental design in some way and improve chances of replicability?

This workshop assumes a little literacy in R and linear mixed-effects models, but I have attempted to make these subjects as accessible as possible, even without knowledge of these topics. If you are interested in gaining hands-on pracical experience with the code, then feel free to download the following .Rmd files with the code, so that you can get a better idea of what each code snippet does and can manipulate them according to your own needs and studies.

```{r, echo = FALSE}
xfun::embed_file('ExerciseDataSimulation.qmd')
```

*LINK TO EXERCISE SHEET, PART 1
*LINK TO EXERCISE SHEET, PART 2
*LINK TO EXERCISE SHEET, PART 3

I hope that this will be fun experience and useful exploration of data simulation, power analysis, statistical modelling and programming, and if you have any questions, big or small, feel free to contact me on chris[dot]mm[dot]cox@gmail.com.