---
title: Part II, Simulation-Based Power Analysis
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#if you don't have the pacman package loaded on your computer, uncomment the next line, install pacman, and load in the required packages
options(repos = "https://cran.r-project.org/")
install.packages('pacman')
#load the required packages:
pacman::p_load(knitr, # rendering of .Rmd file
               lme4, # model specification / estimation
               afex, # anova and deriving p-values from lmer
               broom.mixed, # extracting data from model fits
               faux, # generate correlated values
               tidyverse, # data wrangling and visualisation
               ggridges, #visualisation
               viridis, # color schemes for visualisation
               kableExtra, #helps with knitting the .Rmd file
               cowplot, #visualisation tool to include multiple plots
               ggrain, #visualisation tool for raincloud plots
               dplyr,
               here,
               tidyr,
               readr
               )

set.seed(1234)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE, fig.width=12, fig.height=11, fig.fullwidth=TRUE)

plot_theme <- 
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 18), 
        axis.text.x = element_text(size = 13), 
        axis.title.x = element_text(size = 13), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 13),
        strip.background = element_rect(color="white", fill="white", size=1.5, linetype="solid"))

```

## Cumulative Science and Prior Knowledge
Up until now, we have relied solely on our intuitions about infant looking times based on our experience with conducting infant experiments. This strategy introduces lots of assumptions into the simulation process and creates a multiverse of different parameter settings that have real effects on power analyses. In this section, we consider how to improve our data simulation process by capitalising on cumulative science efforts, such as using effect size estimates from meta-analyses and multi-lab replication studies. 

For the IDS preference effect, for example, we can thank the ManyBabies community for conducting both a multi-lab replication study and a community-augmented meta-analysis on infants’ preference to attend to IDS over ADS (ManyBabies Consortium, 2020; Zettersten, Cox, Bergmann, et al., 2023). By synthesising data across such a wide variety of experimental designs, participants and stimuli, we now have a fairly good estimate of the overall magnitude of the IDS preference effect. Both sources of evidence converge on an effect size of ~0.35 with 95% CI of [0.16; 0.47]. This section delves into the realm of effect sizes and teaches you how to implement a power analysis based on an effect size estimate. 

## Adapting our Simulation Function to Effect Size Data

Let's continue with our IDS preference example and take inspiration from the ManyBabies1 estimate (https://doi.org/10.1177/2515245919900809) to think about how we would simulate data for a new experimental study on the IDS preference effect. Because our hypothetical study still revolves around a within-subjects, between-items study, we can rely on the simulation function from the previous page. All we have to do is to adapt the simulation function so that it suits the new scale of effect sizes.

Let's adapt our simulation function from previous pages to the new scale of effect sizes and call it SimulateEffectSizeData(). Following ManyBabies Consortium (2020), a positive effect size denotes longer looking times to IDS stimuli over ADS stimuli, and an effect size of 0 denotes no preference for either speech style (i.e., similar looking times to ADS and IDS stimuli).

```{r}
# set up the custom data simulation function
SimulateEffectSizeData <- function(
  n_subj = 24,   # number of subjects
  n_ADS  = 6,   # number of ingroup stimuli
  n_IDS =  6,   # number of outgroup stimuli
  mean_intercept = 0,   # grand mean
  mean_slope =  0.35,   # effect of category
  item_varyingintercept =  0.1,   # by-item random intercept sd
  subject_varyingintercept = 0.1,   # by-subject random intercept sd
  subject_varyingslope =  0.2,   # by-subject random slope sd
  rho = 0.2,   # correlation between intercept and slope
  sigma = 0.3) { # residual (standard deviation)

  items <- data.frame(
  #item_id = seq_len(n_ADS + n_IDS),
  Register = rep(c("IDS", "ADS"), c(n_ADS, n_IDS)),
  item_intercept_sd = rnorm(n = n_ADS + n_IDS, mean = 0, sd = item_varyingintercept)
) %>% 
  mutate(SpeechStyle = recode(Register, "ADS" = 0, "IDS" = 1)) %>% 
  mutate(item_id = faux::make_id(nrow(.), "I"))

  # simulate a sample of subjects

# sample from a multivariate random distribution 
  subjects <- faux::rnorm_multi(
  n = n_subj, 
  mu = 0, # means for random effects are always 0
  sd = c(subject_varyingintercept, subject_varyingslope), # set SDs
  r = rho, # set correlation, see ?faux::rnorm_multi
  varnames = c("subject_intercept_sd", "subject_slope_sd")
) %>%
  mutate(subj_id = faux::make_id(nrow(.), "S"))

   ParameterValues <- crossing(subjects, items)  %>%
    mutate(e_si = rnorm(nrow(.), mean = 0, sd = sigma))
  
  # calculate the response variable
  ParameterValues %>%
    mutate(EF = mean_intercept + subject_intercept_sd + item_intercept_sd + (mean_slope + subject_slope_sd) * SpeechStyle + e_si) %>% #sum together overall intercept, varying subject and item intercepts, varying subject slopes, and random error.
  dplyr::select(subj_id, item_id, Register, SpeechStyle, EF)
}
```

```{r}
EffectSizeDataSimulated <- SimulateEffectSizeData()
EffectSizeDataSimulated %>%
    ggplot() + 
  geom_density(aes(EF, fill = Register), alpha = 0.8) +
  geom_vline(xintercept = 0.35, linetype = 3) +
  geom_vline(xintercept = 0, linetype = 3) +
  xlim(c(-5*0.35, 5*0.35)) +
  ggtitle("IDS Preference Effect in Effect Sizes") +
  plot_theme + 
  scale_fill_brewer(palette = "Dark2") + 
  theme(axis.title.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank())
```

```{r, warning = FALSE}
EffectSizeDataSimulated %>%
  group_by(subj_id, Register) %>%
  dplyr::summarise(medLT = mean(EF), .groups = 'drop') %>%
ggplot(aes(x = Register, y = medLT, fill = Register)) + 
  geom_rain(alpha = 0.8, rain.side = "f1x1", id.long.var = "subj_id", point.args.pos = list(position = position_jitter(width = 0.04, height = 0, seed = 42)), line.args.pos = list(position = position_jitter(width = 0.04, height = 0, seed = 42))) + 
  scale_fill_brewer(palette = 'Dark2') +
  ggtitle('Subject-Level Differences across Speech Styles') + 
  xlab("Speech Style") + 
  ylab('Effect Size') + 
  plot_theme
```

## A Linear Mixed Effects Model of Simulated Effect Size Data

Let's think about how we want to run a linear mixed-effects model of the data. For a model with varying intercepts and varying slopes for subject and varying intercepts for item, an appropriate model could involve the following lmer syntax:

```{r}
EffectSizeDataSimulated <- SimulateEffectSizeData()

model <- lmer(EF ~ 1 + SpeechStyle + (1 | item_id) + (1 + SpeechStyle | subj_id), data = EffectSizeDataSimulated)

summary(model)
```

In the following code blocks, we will automatise this process and run the power analysis proper!

## Time to Power-Up the Analysis

Now we have two essential components to perform our simulation-based power analysis: i) a code pipeline to generate data for our research question and ii) a clear idea of how we want to model the data. Now it's time to run the actual power analysis. The way we do this is to specify an effect, run hundreds of models and count the proportion of models that reject the nulle hypothesis. This proportion is an estimate of statistical power for those particular parameter values. 

To simplify the process, we can write a new function that combines a modelling component into our previous SimulateEffectSizeData() function. We will call this function SimulateAndModelEFData(), and we will use broom.mixed::tidy(model) to obtain a dataframe with relevant results from the model and write each model result to a .csv-file.

```{r, warning = FALSE}
# simulate, analyze, and return a table of parameter estimates
SimulateAndModelEFData <- function(...) {
  #simulate EF data function
  dataSimulated <- SimulateEffectSizeData()
  
  #model EF data
  model <- lmer(EF ~ 1 + SpeechStyle + (1 | item_id) + (1 + SpeechStyle | subj_id),
                data = dataSimulated)
  #write to a dataframee
  broom.mixed::tidy(model)
}
```

## Running the Power Analysis

Now we have a function that generates data, runs a model and spits out the results! Now it's time to repeat a few hundred times, so that we can calculate how much power we have with our given parameters. We are going to use map_df() to run the simulation and modelling function 500 times and write it to a .csv file.

```{r, message=FALSE, results='hide',warning = FALSE}
# run simulations and save to a file
n_runs <- 500 # use at least 500 to get stable estimates
simulations <- purrr::map_df(1:n_runs, ~ SimulateAndModelEFData())
write_csv(simulations, "EFsimulations.csv")
```

If it ran correctly, it should have produced a .csv file with model results from each new simulation of data. Let's read in the results and have a look at what they say!

```{r, warning = FALSE}
# read saved simulation data
sims <- read_csv("EFsimulations.csv", show_col_types = FALSE) %>%
  dplyr::select(term, estimate, std.error, p.value)

sims %>% 
  group_by(term) %>%
  dplyr::summarize(
    mean_estimate = mean(estimate),
    mean_se = mean(std.error),
    power = mean(p.value < 0.05),
    .groups = "drop"
  )
```

## Exercises to Check Understanding

### Exercise IV

Now that we have this pipeline set up, it becomes easy to adapt the code to try out different parameter values. Let's explore the effect of repeated measures on power. Try to run a power analysis with each subject receiving two items in each speech style. What happens to the estimate of statistical power?

```{r, message=FALSE, results='hide', warning = FALSE}
#| code-fold: true
#| code-summary: "Show the code"
 
# simulate, analyze, and return a table of parameter estimates
SimulateAndModelEFData <- function(...) {
  #simulate EF data function
  dataSimulated <- SimulateEffectSizeData(n_ADS = 2, n_IDS = 2)
  
  #model EF data
  model <- lmer(EF ~ 1 + SpeechStyle + (1 | item_id) + (1 + SpeechStyle | subj_id),
                data = dataSimulated)
  #write to a dataframee
  broom.mixed::tidy(model)
}

# run simulations and save to a file
n_runs <- 500 # use at least 500 to get stable estimates
simulations <- purrr::map_df(1:n_runs, ~ SimulateAndModelEFData())
write_csv(simulations, "EFsimulations2Stimuli.csv")
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# read saved simulation data
sims <- read_csv("EFsimulations2Stimuli.csv", show_col_types = FALSE) %>%
  dplyr::select(term, estimate, std.error, p.value)

sims %>% 
  group_by(term) %>%
  dplyr::summarize(
    mean_estimate = mean(estimate),
    mean_se = mean(std.error),
    power = mean(p.value < 0.05),
    .groups = "drop"
  )
```

### Exercise V

Let's imagine a scenario where we are interested in the effect of age on IDS preference. We would like to explore the extent to which we can detect a cross-sectional age effect given only two stimulus items per participant. How would adapt the above code to explore this experimental design? 

```{r}
#| code-fold: true
#| code-summary: "Show the code"
 
SimulateEFDataWithAge <- function(
  beta_age = 0.3, #add effect of age
  age_interaction_sd = 0.1, #add some standard deviation to aage effect
  n_subj = 24,   # number of subjects
  n_ADS  = 6,   # number of ingroup stimuli
  n_IDS =  6,   # number of outgroup stimuli
  mean_intercept = 0,   # grand mean
  mean_slope =  0.35,   # effect of category
  item_varyingintercept =  0.1,   # by-item random intercept sd
  subject_varyingintercept = 0.1,   # by-subject random intercept sd
  subject_varyingslope =  0.2,   # by-subject random slope sd
  rho = 0.2,   # correlation between intercept and slope
  sigma = 0.3) { # residual (standard deviation)

  items <- data.frame(
  #item_id = seq_len(n_ADS + n_IDS),
  Register = rep(c("IDS", "ADS"), c(n_ADS, n_IDS)),
  item_intercept_sd = rnorm(n = n_ADS + n_IDS, mean = 0, sd = item_varyingintercept)
) %>% 
  mutate(SpeechStyle = recode(Register, "ADS" = 0, "IDS" = 1)) %>% 
  mutate(item_id = faux::make_id(nrow(.), "I"))

  # simulate a sample of subjects

# sample from a multivariate random distribution 
  subjects <- faux::rnorm_multi(
  n = n_subj, 
  mu = 0, 
  sd = c(subject_varyingintercept, subject_varyingslope, age_interaction_sd), 
  r = rho,
  varnames = c("subject_intercept_sd", "subject_slope_sd", "age_slope_sd")
) %>%
    mutate(subj_id = faux::make_id(nrow(.), "S")) %>%
    mutate(age_subj = runif(n_subj, min = -0.5, max = 0.5))

  ParameterValues <- crossing(subjects, items)  %>%
    mutate(e_si = rnorm(nrow(.), mean = 0, sd = sigma))
  
  # calculate the response variable
  ParameterValues %>%
    mutate(EF = mean_intercept + subject_intercept_sd + item_intercept_sd + (mean_slope + subject_slope_sd) * SpeechStyle + ((beta_age + age_slope_sd) * age_subj * SpeechStyle) + e_si) %>%
  dplyr::select(subj_id, item_id, Register, SpeechStyle, age_subj, EF)
}
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"

DataWithAgeSimulated <- SimulateEFDataWithAge()
DataWithAgeSimulated %>%
ggplot() + 
  geom_point(aes(y = EF, x = age_subj, color = subj_id), alpha = 0.6, size = 1, show.legend = F) + 
  geom_smooth(method = "lm", se = TRUE, formula = y ~ x, aes(y = EF, x = age_subj)) +
  ggtitle("Interaction Effect with Age") +
  xlab("Age (standardised age)") + 
  facet_wrap(~Register) +
  scale_color_manual(values = viridis(n = 27)) +
  plot_theme
```

```{r, warning = FALSE, message=FALSE, results='hide', warning = FALSE}
#| code-fold: true
#| code-summary: "Show the code"
 
# simulate, analyze, and return a table of parameter estimates
SimulateAndModelEFAgeData <- function(...) {
  #simulate EF data function
  dataSimulated <- SimulateEFDataWithAge(n_ADS = 4, n_IDS = 4)
  
  #model EF data
  model <- lmer(EF ~ 1 + SpeechStyle:age_subj + (1 | item_id) + (1 + SpeechStyle | subj_id),
                data = dataSimulated)
  #write to a dataframee
  broom.mixed::tidy(model)
}

# run simulations and save to a file
n_runs <- 500 # use at least 500 to get stable estimates
simulations <- purrr::map_df(1:n_runs, ~ SimulateAndModelEFAgeData())
write_csv(simulations, "EFsimulationsAge.csv")
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"

# read saved simulation data
sims <- read_csv("EFsimulationsAge.csv", show_col_types = FALSE) %>%
  dplyr::select(term, estimate, std.error, p.value)

sims %>% 
  group_by(term) %>%
  dplyr::summarize(
    mean_estimate = mean(estimate),
    mean_se = mean(std.error),
    power = mean(p.value < 0.05),
    .groups = "drop"
  )
```

Now we have a pretty useful pipeline set up that allows us to explore the effects of different parameter values on our ability to detect effects. However, instead of manually varying the parameters one by one, it would be nice if we could set up a grid search to explore values and put the power results into perspective. We will explore how to do this in the next exercise sheet. The code gets slightly more complex, so make sure that you have understood the code that we have written so far before venturing further.
