---
title: Solution to Exercise I
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
#if you don't have the pacman package loaded on your computer, uncomment the next line, install pacman, and load in the required packages
options(repos = "https://cran.r-project.org/")
install.packages('pacman')
#load the required packages:
pacman::p_load(knitr, # rendering of .Rmd file
               lme4, # model specification / estimation
               afex, # anova and deriving p-values from lmer
               broom.mixed, # extracting data from model fits
               faux, # generate correlated values
               tidyverse, # data wrangling and visualisation
               ggridges, #visualisation
               viridis, # color schemes for visualisation
               kableExtra, #helps with knitting the .Rmd file
               cowplot, #visualisation tool to include multiple plots
               ggrain, #visualisation tool for raincloud plots
               dplyr,
               here,
               tidyr
               )

set.seed(1234)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE, fig.width=12, fig.height=11, fig.fullwidth=TRUE)

plot_theme <- 
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 18), 
        axis.text.x = element_text(size = 13), 
        axis.title.x = element_text(size = 13), 
        axis.text.y = element_text(size = 12), 
        axis.title.y = element_text(size = 13),
        strip.background = element_rect(color="white", fill="white", size=1.5, linetype="solid"))

```

```{r}
# set up the custom data simulation function
SimulateLTData <- function(
  n_subj = 24,   # number of subjects
  n_ADS  = 8,   # number of ingroup stimuli
  n_IDS =  8,   # number of outgroup stimuli
  mean_intercept = 7000,   # grand mean
  mean_slope =  2000,   # effect of category
  item_varyingintercept =  200,   # by-item random intercept sd
  subject_varyingintercept = 2000,   # by-subject random intercept sd
  subject_varyingslope =  1000,   # by-subject random slope sd
  rho = 0.2,   # correlation between intercept and slope
  sigma = 500) { # residual (standard deviation)

  items <- data.frame(
  #item_id = seq_len(n_ADS + n_IDS),
  Register = rep(c("IDS", "ADS"), c(n_ADS, n_IDS)),
  O_0i = rnorm(n = n_ADS + n_IDS, mean = 0, sd = item_varyingintercept)) %>% 
  mutate(item_id = faux::make_id(nrow(.), "I")) %>%
  mutate(SpeechStyle = recode(Register, "ADS" = -0.5, "IDS" = +0.5))

  # simulate a sample of subjects

# sample from a multivariate random distribution 
  subjects <- faux::rnorm_multi(
  n = n_subj, 
  mu = 0, # means for random effects are always 0
  sd = c(subject_varyingintercept, subject_varyingslope), # set SDs
  r = rho, # set correlation, see ?faux::rnorm_multi
  varnames = c("T_0s", "T_1s")
) %>%
  mutate(subj_id = faux::make_id(nrow(.), "S"))

  ParameterValues <- crossing(subjects, items)  %>%
    mutate(e_si = rnorm(nrow(.), mean = 0, sd = sigma)) %>%
    dplyr::select(subj_id, item_id, Register, SpeechStyle, everything())
  
  ParameterValues %>%
    mutate(LT = mean_intercept + T_0s + O_0i + (mean_slope + T_1s) * SpeechStyle + e_si) %>%
    mutate(LT = LT + rexp(nrow(.), rate = 0.0003)) %>%
    dplyr::select(subj_id, item_id, Register, SpeechStyle, LT)
}
```

## Solution to Exercise Ia
* How would you adapt our code to generate a dataset with 500 participants and no effect of SpeechStyle?

With our new SimulateLTData() function, the answer here is fairly straightforward! We can simply specify that we want to simulate 500 subjects and want a mean slope of 0, like so: SimulateLTData(n_subj = 500, mean_slope = 0).

```{r}
LTDataSimulated <- SimulateLTData(n_subj = 500, mean_slope = 0)

LTDataSimulated %>%
  group_by(subj_id, Register) %>%
  dplyr::summarise(medLT = mean(LT)) %>%
  ggplot(aes(x = Register, y = medLT, fill = Register)) + 
  geom_rain(alpha = 0.8, rain.side = "f1x1", id.long.var = "subj_id", point.args.pos = list(position = position_jitter(width = 0.04, height = 0, seed = 42)), line.args.pos = list(position = position_jitter(width = 0.04, height = 0, seed = 42))) + 
  scale_fill_brewer(palette = 'Dark2') +
  ggtitle('Looking Time Differences across Speech Styles') + 
  xlab("Speech Style") + 
  ylab('Looking Time (ms)') + 
  scale_color_manual(values = viridis(n = 27)) +
  plot_theme
```

## Solution to Exercise Ib:
Exercise Ib: We might expect the IDS preference effect to change with infant age, such that older infants listen longer to IDS over ADS than younger infants. How would you add a positive interaction effect of (cross-sectional) age as a predictor to the model (hint: it involves randomly sampling age for each child and adding an effect to the simulation code and model)?

The question here involves adding infant age as an interaction effect with SpeechStyle. We will approach this question by modifying the code that simulates subject-level data. Here, we will sample an age variable and pretend that we have standardised age so that its values fall between -0.5 and 0.5. We thus randomly sample and age to assign one age to each subject. We also need to specify a slope value for the influence of subject ag and place it in the start of the function; we will add subject_age = 200. Lastly, we need to change how we sum the values together, so that age has an effect on looking times, but also that the influence of age exerts different effects across the two speech styles.

```{r}
# set up the custom data simulation function
SimulateLTDataWithAge <- function(
  n_subj = 24,   # number of subjects
  n_ADS  = 8,   # number of ingroup stimuli
  n_IDS =  8,   # number of outgroup stimuli
  beta_0 = 7000,   # grand mean
  beta_1 =  15000,   # effect of category
  beta_as = 10000,
  S_as = 50,
  item_sd =  200,   # by-item random intercept sd
  tau_0 = 2000,   # by-subject random intercept sd
  tau_1 =  1000,   # by-subject random slope sd
  rho = 0.2,   # correlation between intercept and slope
  sigma = 500) { # residual (standard deviation)

  items <- data.frame(
  #item_id = seq_len(n_ADS + n_IDS),
  Register = rep(c("IDS", "ADS"), c(n_ADS, n_IDS)),
  O_0i = rnorm(n = n_ADS + n_IDS, mean = 0, sd = item_sd)) %>% 
  mutate(item_id = faux::make_id(nrow(.), "I")) %>%
  mutate(SpeechStyle = recode(Register, "ADS" = 0, "IDS" = 1))

  # simulate a sample of subjects

# sample from a multivariate random distribution 
  subjects <- faux::rnorm_multi(
  n = n_subj, 
  mu = 0, # means for random effects are always 0
  sd = c(tau_0, tau_1, S_as), # set SDs
  r = rho, # set correlation, see ?faux::rnorm_multi
  varnames = c("T_0s", "T_1s", "S_as")
) %>%
  mutate(subj_id = faux::make_id(nrow(.), "S")) %>%
  mutate(age_subj = runif(n_subj, min = -0.5, max = 0.5))

  ParameterValues <- crossing(subjects, items)  %>%
    mutate(e_si = rnorm(nrow(.), mean = 0, sd = sigma)) %>%
    dplyr::select(subj_id, item_id, Register, SpeechStyle, age_subj, S_as, everything())
  
  ParameterValues %>%
    mutate(LT = beta_0 + T_0s + O_0i + (beta_1 + T_1s) * SpeechStyle + ((beta_as + S_as) * age_subj * SpeechStyle) + e_si) %>%
    mutate(LT = LT + rexp(nrow(.), rate = 0.0003)) %>%
    dplyr::select(subj_id, item_id, Register, SpeechStyle, age_subj, LT)
}
```

### Visualisation of the Solution as Sanity Check

```{r}
dat_sim <- SimulateLTDataWithAge()
dat_sim %>%
ggplot() + 
  geom_point(aes(y = LT, x = age_subj, color = subj_id), alpha = 0.6, size = 1) + 
  geom_smooth(method = "lm", se = TRUE, formula = y ~ x, aes(y = LT, x = age_subj)) +
  ggtitle("Adding Age as an Interaction Effect") +
  xlab("Age") + 
  facet_wrap(~SpeechStyle) +
  scale_color_manual(values = viridis(n = 27)) +
  plot_theme
```

